Recurrent Quantification Analysis of Movement Variability in Human-Humanoid Interaction Activities

### 2020-07-02 docs/manuscript/main.tex
completion of corrections-2020-01-14 and updates manuscript2020-07-03

%*******************************************************************************
\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\graphicspath{{../}} %%%REPOSITORY PATH

\title{The Application of Recurrent Quantification Analysis 
to Human-Humanoid Interaction}
\author[1,*]{Miguel Xochicale}
\author[1]{Chris Baber}
\affil[1]{University of Birmingham, 
	School of Engineering, 
	Birmingham, 
	B15 2TT, 
	UK}
\affil[*]{map479@bham.ac.uk}

@@

\begin{abstract}
Human movement variability occurs in motor performance 
across multiple repetitions of a task and such behaviour is an
inherent feature within and between each persons' movement. Quantifying
movement variability is still an open problem, particularly when traditional
methods in time domain and frequency domain fail to detect tiny modulations in
frequency or phase for time series. 
For this work, we hence investigate methodologies from nonlinear dynamics such as 
reconstructed state space (RSS), uniform time-delay embedding (UTDE), 
recurrence plots (RPs) and recurrence quantification analysis (RQA) metrics. 
Particularly, we are interested in the weaknesses and robustness 
of nonlinear dynamics tools from raw and post-processed data of 
wearable inertial sensors (IMUs). 
In the reported experiment, twenty right-handed healthy participants 
imitated simple vertical and horizontal arm movements in normal 
and faster speed from an humanoid robot.
With four window lengths and three levels of smoothed time series, 
we found visual differences in the patterns of RSSs and RPs. 
Then with the use of RQA metrics, we find out that the type of 
movements and the level of smoothness affects those metrics. 
Specifically, entropy values from RQA were well distributed and 
presented variation in all the conditions for time series. 
This work have then the potential to enhance the development of 
better diagnostic tools for various applications in rehabilitation, 
sport science or for new forms of human-robot interaction.
\end{abstract}
%226

@@

\section*{Introduction}
Human movement requires a complex system where not only multiple
joints and limbs are involved for a specific task in a determined environment
but also perception and action of movement affects the physical performance 
\cite{davids2003}. 
In contrast, variability in humanoid movement is usually very small,
as a result of the control system used. This means that, while humanoids 
solve the degrees of freedom problem through join design and algorithms,
humans tend to have more fluid and flexible approach. Consequently, 
one can see much variability in the performance of even the simplest task.
Studies of human motion reveal the possibility to 
estimate features from lower dimension signals to distinguish differences between 
styles of pedalling motion \cite{Quintana-Duque2012, Quintana-Duque2016}, 
gait identification \cite{sama2013, frank2010},
and detection of pathologies \cite{gomezgarcia2014}.
The lower dimension signals from biological systems are time series 
of $1-$dimension in $\mathbb{R}$ which commonly are 
noisy, nonlinear and non-stationary \cite{gomezgarcia2014}.
Hence, nonlinear dynamics can be used to 
objectively quantify variability of lower
dimension signals \cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, 
frank2010, gomezgarcia2014, marwan2011, stergiou2011}.
For instance, Bradley et al. 2015 \cite{bradley2015} reviewed methods for
nonlinear time series analysis based on the appropriate estimation of the 
embedding parameters ($m$ embedding dimension and $\tau$ embedding delay) to 
reconstruct the state space,
where an $n$-dimensional reconstructed state space using $1-$dimensional 
time series,
can preserve the topological properties of an unknown $M$-dimensional 
state space \cite{takens1981}.
Bradley et al. 2015 \cite{bradley2015} also reviewed the use of 
Recurrence Plots (RPs), a graphical representation of a two-dimensional map 
which show black and white dots as recurrences in a given $n$-dimensional system, 
and Recurrence Quantification Analysis (RQA) metrics compute statistics in RPs.
In general, RPs and RQA provide an intuitive meaning of the time series,
for instance, RQA is quantitatively and qualitatively independent of 
embedding dimension (also verified experimentally \cite{iwanski1998}).
However, the estimation of embedding parameters and the selection of the 
right parameters to perform RQA is still an open problem.
Bradley et al. 2015 \cite{bradley2015} pointed out that
there is no general technique that can be used to compute the embedding
parameters since time series are system-dependent which means that
computing the values for embedding parameters may only work for
one purpose (e.g., prediction) and may not work well for another purpose
(e.g., computing dynamical invariants).
Additionally, the methodologies of nonlinear dynamics for 
computing the embedding parameters e.g., autocorrelation, mutual information, 
and nearest neighbour require data which is well 
sampled and with little noise \cite{garland2016}
or require purely deterministic signals \cite{kantz2003}.
Thus, these nonlinear analysis methods for computing the embedding 
parameters can break down
with real-world datasets which have generally different length, 
different values of accuracy and precision (rounding errors due to finite 
precision of the measurement apparatus which include frequency 
acquisition \cite{frank2010}),
and data may be contaminated with different sources of noise
\cite{garland2016}.
It is surprising that even with the previous constraints with regard to 
the quality of data, and the problem with the estimation of embedding parameters,
the use of nonlinear dynamics have proven to be helpful to understand and 
characterise dynamics of time series 
\cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, frank2010,
gomezgarcia2014, gomezgarcia2014, marwan2011, stergiou2011, bradley2015}.

@@

\subsection*{Reconstructed State Spaces}
One of the challenges of the implementation of 
Uniform Time-Delay Embedding to Reconstructed State Spaces is the 
selection of embedding parameters because each time series is unique in 
terms of its structure (modulation of amplitude, frequency and phase) 
\cite{ frank2010, sama2013, bradley2015}.
With that in mind, the problem for this work is not to compute individual 
embedding parameters for each of the time series but to deal with a 
selecting of two parameters that can represent all the time series. 
Our solution for that problem was to compute a sample mean over all 
values in each of the conditions of the time series 
%of Figs~\ref{fig:ts}
for minimum dimension values and 
%Figs~\ref{fig:caoami} 
for minimum delay values as follows.
%, resulting in an average minimum embedding parameters 
%of ($\overline{m_0}=6$, $\overline{\tau_0}=8$).

\subsubsection*{Minimum embedding parameters}
Minimum embedding parameters were firstly computed 
with False Nearest Neighbour and Average Mutual Information.
%for Uniform Time-Delay Embedding (UTDE),
%Reconstructed State Space (RSS), 
%Recurrence Plots (RPs) and 
%Recurrence Quantification Analysis (RQAs).
Figures \ref{fig:caoami}(A) show that minimum embedding values 
for sensor HS01 appear to show more variations as 
interquartile range is near to 1 
with three exceptions (HF sg0, VN sg0, and VF sg1).
Minimum embedding values appear to be constant for sensor RS01 
as interquartile range is near to 0.1 with the exception of 
two axis (HF sg0 and VF sg1).
Additionally, it can be seen in Figs \ref{fig:caoami}(A) a 
decrease of mean values (rhombus) in the box plots
as smoothness of time series increase.
Similarly, the first minimum values of AMI values are 
shown in the box plots of Figs \ref{fig:caoami}(B).
It can be seen that values for HS01 tend to be more spread as the smoothness 
of the time series is increasing 
(see the increase of both mean (rhombus) and interquartile range).
However, AMI values for RS01 do not show such increase in relation with
the increase of smoothness except for HF and VF
(Figs \ref{fig:caoami}(B)).

We also computed an overall minimum embedding parameters that represent 
participants, activities, sensors and levels of smoothness, using a sample 
mean of all values in Figs~\ref{fig:caoami}.
The sample mean for the minimum values of $E_{1}(m)$ from Figs \ref{fig:caoami}(A) 
is $\overline{m}_0=6$ 
and the sample mean for minimum values of AMIs from Figs~\ref{fig:caoami}(B)
is $\overline{\tau}_0=8$.
%Hence, we use the average minimum embedding parameters 
%($\overline{m_0}=6$, $\overline{\tau_0}=8$) to compute 
%Reconstructed State Spaces (RSSs), Recurrence Plots (RPs) and
%Recurrence Quantification Analysis (RQA) metrics for human-humanoid activities.
%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/caoami/pdf/fig2}
	\caption{
	{\bf Box plots of minimum embedding parameters for horizontal and vertical arm movements.} 
	Box plots of (A) minimum embedding dimensions 
	and (B) first minimum AMI values for 
	Horizontal Normal (HN), Horizontal Faster (HF),
	Vertical Normal (VN) and Vertical Faster (VF)
	with sensors attached to participants (HS01) and
	sensor attached to robot (RS01).
	Minimum embedding dimensions are for twenty participants 
	($p01$ to $p20$) with three smoothed signals 
	(sg0: sg0zmuvGyroZ, sg1: sg1zmuvGyroZ and sg2: sg2zmuvGyroZ)
	and window length of 10-sec (500 samples).
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:caoami}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsubsection*{Uniform Time-Delay Embedding}
Uniform time-delay embedding were computed with
embedding parameters ($\overline{m_0}=6$, $\overline{\tau_0}=8$) and 
the first three axis of the rotated data of the PCA are shown 
for the reconstructed state spaces in 
Figs~\ref{fig:rss_aHw10} for horizontal arm movements and 
Figs~\ref{fig:rss_aVw10} for vertical arm movements.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/pdf/fig3}
\caption{
	{\bf RSSs for horizontal arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig:ts}.
	Reconstructed state spaces were computed with 
	embedding parameters $m=6$, $\tau=8$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rss_aHw10}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/pdf/fig4}
    \caption{
	{\bf RSSs for vertical arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig:ts}.
	Reconstructed state spaces were computed with 
	embedding parameters $m=6$, $\tau=8$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rss_aVw10}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

It is easy to observe by eye the differences in each of the
trajectories in the reconstructed state spaces 
(Figs~\ref{fig:rss_aHw10}, \ref{fig:rss_aVw10}), 
however one might be not objective when quantifying those differences 
since those observation might vary from person to person.
With that in mind, we tried to objectively quantify those differences 
using euclidean distances between the origin to each of the points in the 
trajectories, however these results created a suspicious metric, specially 
for trajectories which looked very messy.
With that in mind, we computed Recurrence Plots and 
Recurrence Quantification Analysis to objectively quantify 
the differences in each of the cases of the time series.

\subsection*{Recurrences Plots}
Recurrence Plots (RP) were computed for horizontal arm movements (Fig~\ref{fig:rp_aH}) and 
vertical arm movements (Fig~\ref{fig:rp_aV}) 
using the average embedding parameters ($m=6$, $\tau=8$) 
and an recurrence threshold of $\epsilon=1$.

%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aH}
\caption{
	{\bf RPs for horizontal arm movements.}	
	Recurrence plots were computed with 
	embedding parameters $m=6$, $\tau=8$ and $\epsilon=1$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rp_aH}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aV}
\caption{
	{\bf RPs for vertical arm movements.}	
	Recurrence plots were computed with 
	embedding parameters $m=6$, $\tau=8$ and $\epsilon=1$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:rp_aV}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Recurrence Quantification Analysis} \label{ch6:rqas}
%Considering the RPs for 20 participants performing four activities 
%(HN, HF, VN and VF) with sensors attached to the human and 
%to the humanoid robot (HS01, RS01) and the increase of smoothness 
%of time series (sg0zmuvGyroZ, sg1zmuvGyroZ and sg2zmuvGyroZ),
Similarly to RP, we computed four RQA metrics (REC, DET, RATIO and ENTR) 
with embedding parameters $m=6$, $\tau=8$ and 
recurrence threshold $\epsilon=1$.

\subsubsection*{REC values}
It can be seen in the box plots of Figs~\ref{fig:RQABP}(A) that REC values, 
representing the \% of black dots in the RPs, 
are more spread for HN and VN movements (higher interquartile range) 
than HF and VF movements (lower interquartile range) for HS01 sensor. 
In contrast, REC values for RS01 sensor present little variation 
(interquartile range of 0.01).
With regard to the increase of smoothness of time series 
(sg0, sg1 and sg2), REC values present little 
variation as the smoothness is increasing for time series from HS01 
(changes of mean values (rhombus)) while REC values are more affected with 
the smoothness for data from RS01 
(see the incremental changes of mean values (rhombus)).
%See Figs~\ref{fig:rec_aH} and \ref{fig:rec_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual REC values for each participant.

\subsubsection*{DET values}
Figs \ref{fig:RQABP}(B) illustrate DET values, 
representing predictability and organisation of the RPs, 
which change very little (interquartile range is around 0.1) 
for type of movement, level of smoothness or type of sensor.
There is also increase of DET values as the smoothness of the signal increase 
(see the incremental changes of mean values (rhombus)).
It can be noted that the interquartile range for faster movements
(HF and VF) with no smoothing (sg0) is lower than the other
levels of smoothness (sg1 and sg2).
%See Figs~\ref{fig:det_aH} and \ref{fig:det_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual DET values for each participant.

\subsubsection*{RATIO values}
Figs \ref{fig:RQABP}(C) present RATIO values, representing dynamic transitions, 
for horizontal and vertical movements.
It can be seen that RATIO values for HS01 sensor vary less 
for HN movements (interquartile range around 2)
than HF movements (interquartile range around 5).
%which is a similar behaviour of RATIO values for oRS01 sensor.
For faster movements (HF and VF), it can be noted a decrease of 
RATIO values as the smoothness of the time series is increasing (rhombus).
For normal movements (HN and VN), the decrease of RATIO values
is less evident than faster movements (see rhombus).
%See Figs~\ref{fig:ratio_aH} and \ref{fig:ratio_aV} in appendix 
%\ref{appendix:e:ep} 
%for more details about individual RATIO values for each participant.

\subsubsection*{ENTR values}
Figs \ref{fig:RQABP}(D) show ENTR values, Shannon entropy values 
which represent the complexity of 
the structure the time series, 
for type of movement, level of smoothness or type of sensor.
ENTR values for HS01 sensor show a higher variation  
(interquartile range around 0.5)
than ENTR values for RS01 sensor which appear 
to be more constant (interquartile range 0.1).
It can also be noted the increase of smoothness of time series 
also is presented with an increase of ENTR values (see gray rhombus).
%See Figs~\ref{fig:entr_aH} and \ref{fig:entr_aV} in appendix
%\ref{appendix:e:ep}
%for more details about individual ENTR values for each participant.

%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/rqa/pdf/rqa-bp}
    \caption
	[Box plots for RQA values]{
	{\bf Box plots for RQA values.}
	Box plots of (A) REC, (B) DET, (C) RATIO, and (D) ENTR values 
	for 20 participants performing HN, HF, VN and VF movements
	with sensors HS01, RS01 and three smoothed-normalised  
	time series (sg0, sg1 and sg2).
	RQA values were computed with 
	embedding parameters $m=6$, $\tau=8$ and $\epsilon=1$.
	Code and data to reproduce the figure is available from \cite{srep2019}.
        }
    \label{fig:RQABP}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

>>>>

%*******************************************************************************
\documentclass[fleqn,10pt]{wlscirep}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\graphicspath{{../}} %goes to path: docs/

\title{Recurrent Quantification Analysis of Movement Variability in Human-Humanoid Interaction Activities}
\author[1,*]{Miguel Xochicale}
\author[1]{Chris Baber}
\affil[1]{University of Birmingham, 
	School of Engineering, 
	Birmingham, 
	B15 2TT, 
	UK}
\affil[*]{map479@bham.ac.uk}

@@

\begin{abstract}
Human movement variability occurs in motor performance 
across multiple repetitions of a task and such behaviour is an
inherent feature within and between each persons' movement. Quantifying
movement variability is still an open problem, particularly 
when methods in time domain, frequency domain or nonlinear dynamics 
can break down due to the real-world time-series datasets. 
For this work, we therefore investigate nonlinear dynamics methods 
such as reconstructed state space (RSSs), uniform time-delay embedding, 
recurrence plots (RPs) and recurrence quantification analysis metrics (RQAs)
with real-world time-series data.
Particularly, we are interested in the weaknesses and robustness 
of nonlinear dynamics tools with the use of raw and post-processed 
data of wearable inertial sensors. 
That said, twenty right-handed healthy participants 
imitated simple vertical and horizontal arm movements in normal 
and faster velocity from an humanoid robot in order to have 
four window lengths and three levels of smoothed time-series data,
to then found visual differences in the patterns with RSSs and RPs
and particulatly the computed differences with RQA metrics
that help us to quantify activities, types of sensors, windows lenghts 
and level of smoothness. 
Specifically, we can conclude that RQA ENTR, Shannon Entropy, 
can lead to interesting results on the quantification of 
movement variablity for participants of different ages, 
state of health and anthropomorphic features to 
then enhance the development of 
better diagnostic tools for applications in rehabilitation, 
sport science or for new forms of human-humanoid interaction.
\end{abstract}

@@

\section*{Introduction}
Human movement requires a complex system where not only multiple
joints and limbs are involved for a specific task in a determined environment
but also perception and action of movement that affects such 
physical performance \cite{davids2003}. 
In contrast, variability in humanoid movement is usually very small,
as a result of mechanical and dynamic capabilities \cite{gouaillier2009}. 
This means that, while humanoids 
solve the degrees of freedom problem through join design and algorithms,
humans tend to have more fluid and flexible approach. Consequently, 
one can see much variability in human movement performance of even the simplest task.
Studies of human motion reveal the possibility to 
estimate features from lower dimension signals to distinguish differences between 
styles of pedalling motion \cite{Quintana-Duque2012, Quintana-Duque2016}, 
gait identification \cite{sama2013, frank2010},
and detection of pathologies \cite{gomezgarcia2014}.
The lower dimension signals from biological systems are time series 
of $1-$dimension in $\mathbb{R}$ which commonly are 
noisy, nonlinear and non-stationary \cite{gomezgarcia2014}.
Hence, nonlinear dynamics can be used to 
objectively quantify variability of lower
dimension signals \cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, 
frank2010, gomezgarcia2014, marwan2011, stergiou2011}.
For instance, Bradley et al. 2015 \cite{bradley2015} reviewed methods for
nonlinear time series analysis based on the estimation of the 
embedding parameters ($m$ embedding dimension and $\tau$ embedding delay) to 
reconstruct the state space,
where an $n$-dimensional reconstructed state space using $1-$dimensional 
time series,
can preserve the topological properties of an unknown $M$-dimensional 
state space \cite{takens1981}.
Similarly, Bradley et al. 2015 \cite{bradley2015} reviewed the use of 
Recurrence Plots (RPs), a graphical representation of a two-dimensional map 
which show black and white dots as recurrences in a given $n$-dimensional system, 
and Recurrence Quantification Analysis (RQA) metrics that compute statistics in RPs.
In general, RPs and RQA provide an intuitive meaning of the time series,
for instance, RQA is quantitatively and qualitatively independent of 
embedding dimension (also verified experimentally \cite{iwanski1998}).
However, the estimation of embedding parameters and the selection of the 
right parameters to perform RQA is still an open problem.
Bradley et al. 2015 \cite{bradley2015} pointed out that
there is no general technique that can be used to compute the embedding
parameters since time series are system-dependent which means that
computing the values for embedding parameters may only work for
one purpose (e.g., prediction) and may not work well for another purpose
(e.g., computing dynamical invariants).
Additionally, methods of nonlinear dynamics for 
computing embedding parameters e.g., autocorrelation, mutual information, 
and nearest neighbour require data which is well 
sampled and with little noise \cite{garland2016}
or require signals that are purely deterministic \cite{kantz2003}.
Thus, these methods of nonlinear analysis for computing the embedding 
parameters can break down
with real-world datasets which have generally different length, 
different values of accuracy and precision (rounding errors due to finite 
precision of the measurement apparatus which include frequency 
acquisition \cite{frank2010}),
and contaminated data with different sources of noise
\cite{garland2016}.
It is then surprising that even with the previous constraints with regard to 
the quality of data, and the problem with the estimation of embedding parameters,
the use of nonlinear dynamics have proven to be helpful to understand and 
to characterise dynamics of time series 
\cite{Quintana-Duque2012, Quintana-Duque2016, sama2013, frank2010,
gomezgarcia2014, gomezgarcia2014, marwan2011, stergiou2011, bradley2015}.

@@

\subsection*{Reconstructed State Spaces}
One of the challenges of the implementation of 
uniform time-delay embedding to reconstruct the state spaces is the 
selection of embedding parameters because each time series is unique in 
terms of its structure (e.g. modulation of amplitude, frequency and phase) 
\cite{ frank2010, sama2013, bradley2015}.
With that in mind, the problem for this work is not to compute individual 
embedding parameters for each of the time series but to deal with a 
selection of two parameters that can represent all the time series. 
Our solution for such problem, as explained in below sections, 
was to compute a sample mean over all 
values in each of the conditions of the time series 
%of Figs~\ref{fig:ts}
for minimum dimension values and 
%Figs~\ref{fig:caoami} 
for minimum delay values.
%, resulting in an average minimum embedding parameters 
%of ($\overline{m_0}=6$, $\overline{\tau_0}=8$).

\subsubsection*{Minimum embedding parameters}
Minimum embedding parameters were firstly computed 
with the methods of False Nearest Neighbour and Average Mutual Information.
%for Uniform Time-Delay Embedding (UTDE),
%Reconstructed State Space (RSS), 
%Recurrence Plots (RPs) and 
%Recurrence Quantification Analysis (RQAs).
Figures \ref{fig:cao_ami}(A) show box plots for the minimum embedding values 
for sensors HS01 and RS01. Minimum embedding vaues for HS01 presents more 
variations than RS01 as interquartile range for HS01 is near to 1 (with the exceptions of HF sg0, VN sg0, and VF sg1)
whereas interquartile range for RS01 is near to 0.1 (with the exception of two axis HF sg0 and VF sg1).
Additionally, Figures \ref{fig:cao_ami}(A) show
a decrease of mean values (rhombus) in the box plots
as smoothness of time series increase (see sg0, sg1, sg2).
Figures \ref{fig:cao_ami}(B) show box plots for the first minimum values of AMI values.
It can be seen that values for HS01 tend to be more spread as the smoothness 
of the time series is increasing 
(see the increase of both mean (rhombus) and interquartile range).
However, AMI values for RS01 do not show such increase in relation with
the increase of smoothness except for HF and VF
(Figs \ref{fig:cao_ami}(B)).

We also computed a sample mean for an overall minimum embedding parameters that 
represent all participants, activities, sensors and levels of smoothness.
The sample mean for the minimum values of $E_{1}(m)$ from Figs \ref{fig:cao_ami}(A) 
is $\overline{m}_0=6$ 
and the sample mean for minimum values of AMIs from Figs~\ref{fig:cao_ami}(B)
is $\overline{\tau}_0=8$.
%Hence, we use the average minimum embedding parameters 
%($\overline{m_0}=6$, $\overline{\tau_0}=8$) to compute 
%Reconstructed State Spaces (RSSs), Recurrence Plots (RPs) and
%Recurrence Quantification Analysis (RQA) metrics for human-humanoid activities.
%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/caoami/pdf/fig2}
	\caption{
	{\bf Box plots of minimum embedding parameters.} 
	Box plots of (A) minimum embedding dimensions 
	and (B) first minimum AMI values for 
	Horizontal Normal (HN), Horizontal Faster (HF),
	Vertical Normal (VN) and Vertical Faster (VF)
	with sensors attached to participants (HS01) and
	sensor attached to robot (RS01).
	Minimum embedding parameters are for twenty participants 
	($p01$ to $p20$) with three smoothed signals 
	(sg0: sg0zmuvGyroZ, sg1: sg1zmuvGyroZ and sg2: sg2zmuvGyroZ)
	and window length of 10-sec (500 samples).
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:cao_ami}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsubsection*{Uniform Time-Delay Embedding}
Uniform time-delay embedding were computed with the overall 
embedding parameters ($\overline{m_0}=6$, $\overline{\tau_0}=8$) and 
the first three axis of the rotated data of the PCA are shown 
for the reconstructed state spaces of horizontal arm movements
(Figs~\ref{fig:rss_aHw10}) and vertical arm movements 
(Figs~\ref{fig:rss_aVw10}).
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/pdf/fig3}
\caption{
	{\bf RSSs for horizontal arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig:ts}.
	Reconstructed state spaces were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$.
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:rss_aHw10}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/pdf/fig4}
    \caption{
	{\bf RSSs for vertical arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig:ts}.
	Reconstructed state spaces were computed with 
	embedding parameters
	$\overline{m}_0=6$, $\overline{\tau}_0=8$.
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:rss_aVw10}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

One can observe by eye the differences in each of the
trajectories of the reconstructed state spaces 
(Figs~\ref{fig:rss_aHw10} and \ref{fig:rss_aVw10}), 
however an objective quantification is required.
That being said, we tried to quantify those differences 
using euclidean distances between the origin 
to each of the points in their trajectories, 
which results did not capture the dynamics of 
the trajectories specially 
for those that looked very messy.
Such ambiguities lead us to consider the use of 
Recurrence Plots and Recurrence Quantification Analysis 
to objectively quantify dynamics the capture activities 
using time series.

\subsection*{Recurrences Plots}
Recurrence Plots (RP) were computed for horizontal arm movements (Fig~\ref{fig:rp_aH}) and 
vertical arm movements (Fig~\ref{fig:rp_aV}) 
using the average embedding parameters 
($\overline{m}_0=6$, $\overline{\tau}_0=8$) 
and an recurrence threshold of $\epsilon=1$.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aH}
\caption{
	{\bf RPs for horizontal arm movements.}	
	Recurrence plots were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$, and $\epsilon=1$.
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:rp_aH}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rps/pdf/rp_aV}
\caption{
	{\bf RPs for vertical arm movements.}	
	Recurrence plots were computed with 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$, and $\epsilon=1$.
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:rp_aV}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

\subsection*{Recurrence Quantification Analysis} \label{ch6:rqas}
Four Recurrence Quantification Analysis (RQA) metrics (REC, DET, RATIO and ENTR) 
were also computed with $\overline{m}_0=6$, $\overline{\tau}_0=8$, and $\epsilon=1$.

\subsubsection*{REC values}
Figs~\ref{fig:RQABP}(A) show box plots of REC values 
that represent the \% of black dots in RPs.
It can be noted that REC values are more spread 
for HN and VN movements (higher interquartile range) than 
for HF and VF movements (lower interquartile range) for HS01 sensor. 
In contrast, REC values for RS01 sensor present little variation 
(interquartile range of 0.01).
Regarding the increase of smoothness for time series 
(sg0, sg1 and sg2), REC values present little 
variation as the smoothness is increasing for time series from HS01 
(changes of mean values (rhombus)) whereas REC values are 
more affected with the smoothness for data from RS01 
(see the incremental changes of mean values (rhombus)).
%See Figs~\ref{fig:rec_aH} and \ref{fig:rec_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual REC values for each participant.

\subsubsection*{DET values}
Figs \ref{fig:RQABP}(B) illustrate DET values 
that represent the predictability and organisation of RPs.
Generally, it can be noted little change of DET values 
(interquartile range is around 0.1) 
for type of movement, type of sensor 
but the increase of DET values as the smoothness of 
the signal increase 
(see the incremental changes of mean values (rhombus)).
However, the interquartile range for faster movements
(HF and VF) with no smoothing (sg0) is lower than the other
levels of smoothness (sg1 and sg2).
%See Figs~\ref{fig:det_aH} and \ref{fig:det_aV} in appendix \ref{appendix:e:ep} 
%for more details about individual DET values for each participant.

\subsubsection*{RATIO values}
Figs \ref{fig:RQABP}(C) illustrate RATIO values that represent the 
dynamic transitions in RPs. 
Generally, it can be seen that RATIO values for HS01 sensor vary less 
for HN movements (interquartile range around 2)
than HF movements (interquartile range around 5).
For faster movements (HF and VF), it can be noted a decrease of 
the mean values the smoothness of the time series 
is increasing (rhombus). However, for normal movements (HN and VN), 
the decrease of mean values is less evident than 
faster movements (rhombus).
%See Figs~\ref{fig:ratio_aH} and \ref{fig:ratio_aV} in appendix 
%\ref{appendix:e:ep} 
%for more details about individual RATIO values for each participant.

\subsubsection*{ENTR values}
Figs \ref{fig:RQABP}(D) show ENTR values that represent 
Shannon entropy values in RPs. 
Generally, 
ENTR values for HS01 sensor show a higher variation  
(interquartile range around 0.5)
than ENTR values for RS01 sensor 
(interquartile range 0.1).
It can also be noted the increase of mean values 
for ENTR values as the smoothness of time series 
increase (rhombus) for type of movement, 
type of sensor and level of smoothness.  
%See Figs~\ref{fig:entr_aH} and \ref{fig:entr_aV} in appendix
%\ref{appendix:e:ep}
%for more details about individual ENTR values for each participant.

%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/rqa/output/rqa-bp}
    \caption
	[Box plots for RQA values]{
	{\bf Box plots for RQA metrics.}
	RQA metrics for (A) REC, (B) DET, (C) RATIO, and (D) ENTR of 
	20 participants performing HN, HF, VN and VF movements
	with sensors HS01, RS01 and three smoothed-normalised  
	time series (sg0, sg1 and sg2).
	RQA values were computed with 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$, and $\epsilon=1$. 
	Code and data to reproduce the figure is available in \cite{srep2020}.
        }
    \label{fig:RQABP}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

%\subsection*{RQA metrics with fixed parameters}
%Considering that RQA metrics were computed with fixed embedding parameters 
%($m=6$ and $\tau=8$) and recurrence thresholds ($\epsilon=1$), we found 
%the following. REC values, which represents the \% of black points in the RPs, 
%were more affected with and increase in normal speed movements (HN and VN) 
%than faster movements (HF and VF) for the sensor attached to the participants 
%(HS01). Such decrease of REC values from normal speed to faster speed 
%movements is also presented in data from sensor attached to the robot (RS01), 
%and little can be said with regard to the dynamics of the time series coming 
%from RS01 (Fig \ref{fig:RQABP}A).
%Similarly, DET values, representing predictability and 
%organisation in the RPs, present little variation in the any of the time 
%series where little can be said (Fig \ref{fig:RQABP}B).
%In contrast, RATIO values, which represent 
%dynamic transitions, were more variable for faster movements (HF and VF) 
%than normal speed movements (HN and VN) with sensors attached to the 
%participants (HS01). For data coming from sensors attached to the robot 
%(RS01), RATIO values from horizontal movements (HN, HF) appear to vary 
%more than values coming from vertical movmentes (VN, VF) 
%(Fig \ref{fig:RQABP}C).
%With that, it can be said that RATIO values can represent a bit better
%than REC or DET metrics for the variability of imitation activities in 
%each of the conditions for time series.
%Similarly, ENTR values for HN were higher than values for HF
%and ENTR values varied more for sensor attached to participants 
%than ENTR values for sensors of the robot. It is evidently that 
%the higher the entropy the more complex the dynamics are, 
%however, ENTR values for HN appear a bit higher than HF values, 
%for which we believe this happens because of the structure the time series
%which appear more complex for HN than  HF movements which presented a 
%more consistence repetition (Fig \ref{fig:RQABP}D).

\subsection*{3D RQA ENTR}
To show how that the selection of recurrence threshold affects RQA
values, 
we computed ENTR values of RQA metrics for different embedding parameters
$ \{ m \in \mathbb{R} | 1 \le m \le 10  \} $,
$  \{ \tau \in \mathbb{R} | 1 \le \tau \le 10  \} $ incrementing by one,
with the consideration of three recurrence thresholds $\epsilon=1, 2, 3$ 
and three levels of smoothness (sg0, sg1, sg2). 
For instance, 
Figures \ref{fig:RQA-IND} show the increase of recurrence threshold is 
associated to the increase of ENTR values in any of the levels 
of smoothness (see values of the ENTR bar).
Similarly, it can be noted that the increase of level of 
smoothness (sg0, sg1 and sg2) is associated with the increase 
of ENTR values of the 3D surface plots.
\begin{figure}
\includegraphics[width=1.0\textwidth]{figures/rqa/output/rqa-epsilons}
    \caption
